---
title: "Washington Education Results"
subtitle: "District outcomes, resources, equity, and geography"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    self_contained: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "100%"
)

library(dplyr)
library(htmltools)
library(knitr)
library(leaflet)
library(scales)
library(sf)

find_project_root <- function(start_dir = getwd()) {
  current_dir <- normalizePath(start_dir, winslash = "/", mustWork = TRUE)

  repeat {
    expected_data_path <- file.path(
      current_dir,
      "Data",
      "converted",
      "1. Data for Student Performance Paper 2015 2025.rds"
    )

    if (file.exists(expected_data_path)) {
      return(current_dir)
    }

    parent_dir <- dirname(current_dir)
    if (identical(parent_dir, current_dir)) {
      stop("Could not find the project root containing Data/converted.", call. = FALSE)
    }

    current_dir <- parent_dir
  }
}

project_root <- find_project_root()
figures_dir <- file.path(project_root, "Outputs", "figures")
data_path <- file.path(
  project_root,
  "Data",
  "converted",
  "1. Data for Student Performance Paper 2015 2025.rds"
)
boundary_dir <- file.path(project_root, "Data", "geo", "wa_unified_school_districts_2025")
shape_path <- list.files(boundary_dir, pattern = "\\.shp$", full.names = TRUE)[1]

score_data <- readRDS(data_path) |>
  mutate(
    nces_lea = as.character(nces_lea),
    math_score = metMATH_pct * 100,
    ela_score = metELA_pct * 100
  ) |>
  group_by(year, nces_lea, name, county) |>
  summarize(
    math_score = mean(math_score, na.rm = TRUE),
    ela_score = mean(ela_score, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    math_score = if_else(is.nan(math_score), NA_real_, math_score),
    ela_score = if_else(is.nan(ela_score), NA_real_, ela_score)
  )

latest_year <- max(score_data$year, na.rm = TRUE)

map_data <- st_read(shape_path, quiet = TRUE) |>
  st_transform(4326) |>
  mutate(nces_lea = GEOID) |>
  left_join(score_data |> filter(year == latest_year), by = "nces_lea")

district_name <- coalesce(map_data$name, map_data$NAME, "Unknown district")

score_map <- leaflet(map_data, options = leafletOptions(zoomControl = TRUE)) |>
  addProviderTiles(providers$CartoDB.Positron)

score_palette <- colorNumeric(
  palette = rev(viridisLite::viridis(256)),
  domain = c(5, 85),
  na.color = "#d7dde3"
)

for (subject in c("Math", "ELA")) {
  score_column <- if (subject == "Math") "math_score" else "ela_score"
  values <- map_data[[score_column]]
  labels <- HTML(paste0(
    "<strong>", htmlEscape(district_name), "</strong><br>",
    subject, " requirement rate: ",
    ifelse(is.na(values), "No data", paste0(number(values, accuracy = 0.1), "%")),
    "<br>Year: ", latest_year
  ))

  score_map <- score_map |>
    addPolygons(
      group = subject,
      fillColor = score_palette(values),
      fillOpacity = 0.82,
      color = "#ffffff",
      weight = 0.7,
      label = labels,
      highlightOptions = highlightOptions(
        weight = 2,
        color = "#243b53",
        fillOpacity = 0.92,
        bringToFront = TRUE
      )
    )
}

score_map <- score_map |>
  addLayersControl(
    overlayGroups = c("Math", "ELA"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  hideGroup("ELA") |>
  setView(lng = -120.75, lat = 47.4, zoom = 7)

median_map <- leaflet(map_data, options = leafletOptions(zoomControl = TRUE)) |>
  addProviderTiles(providers$CartoDB.Positron)

median_palette <- colorNumeric(
  palette = colorRampPalette(c("#d65f5f", "#f7f7f7", "#3b82a0"))(256),
  domain = c(-50, 50),
  na.color = "#d7dde3"
)

for (subject in c("Math", "ELA")) {
  score_column <- if (subject == "Math") "math_score" else "ela_score"
  values <- map_data[[score_column]]
  subject_median <- median(values, na.rm = TRUE)
  distance <- values - subject_median
  labels <- HTML(paste0(
    "<strong>", htmlEscape(district_name), "</strong><br>",
    subject, " requirement rate: ",
    ifelse(is.na(values), "No data", paste0(number(values, accuracy = 0.1), "%")),
    "<br>Median district rate: ", number(subject_median, accuracy = 0.1), "%",
    "<br>Distance from median: ",
    ifelse(is.na(distance), "No data", paste0(number(distance, accuracy = 0.1, style_positive = "plus"), " pts")),
    "<br>Year: ", latest_year
  ))

  median_map <- median_map |>
    addPolygons(
      group = subject,
      fillColor = median_palette(distance),
      fillOpacity = 0.82,
      color = "#ffffff",
      weight = 0.7,
      label = labels,
      highlightOptions = highlightOptions(
        weight = 2,
        color = "#243b53",
        fillOpacity = 0.92,
        bringToFront = TRUE
      )
    )
}

median_map <- median_map |>
  addLayersControl(
    overlayGroups = c("Math", "ELA"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  hideGroup("ELA") |>
  setView(lng = -120.75, lat = 47.4, zoom = 7)

figure <- function(filename) {
  include_graphics(file.path(figures_dir, filename))
}
```

<style>
body {
  color: #2f3437;
}

h1, h2, h3 {
  color: #243b53;
}

.section-intro {
  color: #697077;
  font-size: 16px;
  margin-bottom: 20px;
}

.chart-block {
  margin-bottom: 34px;
  padding-bottom: 24px;
  border-bottom: 1px solid #d9dee5;
}

.leaflet {
  margin-bottom: 22px;
}
</style>

# Overview

This standalone report presents Washington district-level outcomes, resources,
equity comparisons, and interactive geography. Dollar values shown in the
results are CPI-adjusted to 2015. All comparisons are observational and
descriptive; they do not establish causal effects.

# Outcomes

<p class="section-intro">Statewide Math and ELA requirement rates across the study period.</p>

```{r outcomes-chart}
figure("wa_test_scores_over_time.png")
```

# Resources

<p class="section-intro">Teacher salary and instructional spending compared with composite student outcomes.</p>

## Teacher salaries and outcomes

```{r salary-chart}
figure("wa_salary_vs_scores_scatter.png")
```

## Instructional spending and outcomes

```{r spending-outcomes-chart}
figure("viz_spend_final.png")
```

# Equity

<p class="section-intro">Resources, demographics, and outcomes compared across district poverty levels.</p>

## Spending by poverty quartile

```{r poverty-spending-chart}
figure("viz_spend_by_poverty.png")
```

## Student demographics

```{r demographics-chart}
figure("viz4_demo_comparison.png")
```

## Pay gap and learning gap

```{r pay-learning-chart}
figure("viz4_wealth_poverty_profiles.png")
```

# Explore Maps {.tabset}

The standalone maps show the latest available year, `r latest_year`. Use the
layer control to switch between Math and ELA.

## Requirement Rates

```{r requirement-map}
score_map
```

## Median Comparison

```{r median-map}
median_map
```
